---
title: "Final Project Basic Outline"
author: "Cael Walter"
output: html_notebook
---

This is an outline for your final project. You can follow this outline, or you can modify it as you see fit. Do what works best for your project. 

## Research Question 

State your general specific question and why it is interesting. 
Research Question: How do the Pittsburgh Pirates and the Chicago Cubs compare in terms of performance metrics over the past decade?

Why This Question is Interesting: The rivalry between the Pittsburgh Pirates and the Chicago Cubs is a significant one in MLB, with both teams having experienced periods of success and struggle. Analyzing their performance metrics over the past decade can provide insights into team strategies, player contributions, and overall competitiveness.


## Data Processing 


#### Data Intake 

Include R code and a written explanation to import your **TWO** data sets.
```{r}
# Load necessary libraries
library(tidyverse)
library(rvest)

# Define function to scrape boxscore data
scrape_boxscore <- function(url) {
  page <- read_html(url)
  
  teams <- page %>%
    html_nodes(".team-name a") %>%
    html_text()
  
  scores <- page %>%
    html_nodes(".score") %>%
    html_text() %>%
    as.numeric()
  
  data.frame(
    Team1 = teams[1],
    Score1 = scores[1],
    Team2 = teams[2],
    Score2 = scores[2],
    Date = Sys.Date() # Placeholder for date extraction
  )
}

# Scrape the four provided boxscore links
page <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401569056#:~:text=Play%2Dby%2DPlay-,CHC,0,-WIN")
tableList <- page %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

box_table1 <- tableList[[2]]


head(box_table1)


page <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401472023#:~:text=Play%2Dby%2DPlay-,PIT,0,-WIN")
tableList <- page %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

box_table2 <- tableList[[2]]


head(box_table2)

page <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401354772#:~:text=Play%2Dby%2DPlay-,PIT,0,-WIN")
tableList <- page %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

box_table3 <- tableList[[2]]


head(box_table3)

page <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401227563#:~:text=Play%2Dby%2DPlay-,PIT,2,-WIN")
tableList <- page %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

box_table4 <- tableList[[2]]


head(box_table4)
```

#### Data Wrangling 

Include R code and written explanation for wangling your data (you can make multiple wrangled data sets). 
```{r}
# Load necessary libraries
library(tidyverse)
library(rvest)
library(reshape2)

# Define a function to process each table and extract inning scores
process_boxscore <- function(table, team) {
  # Convert relevant columns to numeric and rename innings
  innings <- table %>%
    select(1:9) %>%
    rename_with(~paste0("Inning", 1:9)) %>%
    mutate(across(everything(), as.numeric))
  
  # Extract total runs, hits, and errors
  scores <- table %>%
    select(R, H, E) %>%
    mutate(across(everything(), as.numeric))
  
  # Combine innings scores with total scores
  combined <- bind_cols(innings, scores) %>%
    mutate(Team = team, GameID = NA)  # Add Team and placeholder for GameID
  
  return(combined)
}

# Scrape the boxscore data from the given URLs
page1 <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401569056")
tableList1 <- page1 %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)
box_table1 <- tableList1[[2]]

page2 <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401472023")
tableList2 <- page2 %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)
box_table2 <- tableList2[[2]]

page3 <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401354772")
tableList3 <- page3 %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)
box_table3 <- tableList3[[2]]

page4 <- read_html("https://www.espn.com/mlb/boxscore/_/gameId/401227563")
tableList4 <- page4 %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)
box_table4 <- tableList4[[2]]

# Process each boxscore table and add team information
boxscore1 <- process_boxscore(box_table1, "CHC")
boxscore2 <- process_boxscore(box_table2, "PIT")
boxscore3 <- process_boxscore(box_table3, "PIT")
boxscore4 <- process_boxscore(box_table4, "PIT")

# Add GameID column for clarity
boxscore1$GameID <- "May 11,2024"
boxscore2$GameID <- "June 13,2023"
boxscore3$GameID <- "May 16, 2022"
boxscore4$GameID <- "May 9, 2021"

# Combine all processed boxscore tables into one data frame
all_boxscores <- bind_rows(boxscore1, boxscore2, boxscore3, boxscore4)

# Ensure the Team column is present and properly formatted
all_boxscores <- all_boxscores %>%
  mutate(Team = as.factor(Team))

# Preview the combined data frame
head(all_boxscores)





```



## Data Visualization

Include R code and written explanation for your data visualization(s) with at least 3 variables. You must have **at least** one graph. You may have more than one (I would encourage you to have more than one). 
```{r}
# Load necessary library for visualization
library(ggplot2)
library(reshape2)

# Melt the data for better visualization
melted_boxscores <- melt(all_boxscores, id.vars = c("GameID", "Team"), variable.name = "Inning", value.name = "Score")

# Example Visualization 1: Inning-by-Inning Scores for Each Team
ggplot(melted_boxscores %>% filter(grepl("^Inning", Inning)), aes(x = Inning, y = Score, color = Team, group = Team)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ GameID) +
  labs(title = "Inning-by-Inning Scores for Each Team", x = "Inning", y = "Score")
```

```{r}
# Example Visualization 2: Total Runs by Team and Game
ggplot(all_boxscores, aes(x = GameID, y = R, fill = Team)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Runs by Team and Game", x = "Game ID", y = "Total Runs")
```

```{r}
# Example Visualization 3: Hits by Team and Game
ggplot(all_boxscores, aes(x = GameID, y = H, fill = Team)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hits by Team and Game", x = "Game ID", y = "Hits")

```



## Conclusion

Answer your research question using the data sets and visualizations you created. 

From the analysis and visualizations, we can conclude that:

PIT (Pittsburgh Pirates):
Consistently performed well across the games with a significant number of hits and runs.
Showed a strong distribution of scores, particularly in the middle innings (e.g., 4th and 6th innings in the May 11, 2024 game).

CHC (Chicago Cubs):
Had a notable performance in game June 13, 2023, scoring 10 runs and having a considerable number of hits.
Struggled in the other games where they failed to score any runs.

Overall, PIT demonstrated consistent offensive strength across the analyzed games, whereas CHC had a standout performance in one game but lacked consistency across the others. This analysis highlights the varying performances of the teams and provides insight into their scoring patterns and overall effectiveness in the recent matches.